---
title: "Seminar: MLE with S&P 500 Returns"
subtitle: "Estimating mean return and volatility using R"
author: ""
date: ""
output:
  html_document:
    toc: true
    toc_float: true
    theme: null
    highlight: tango
    code_folding: show
    df_print: paged
---

<style>
:root{--bg:#f6f7f8;--surface:#ffffff;--ink:#20252b;--muted:#5f6872;--accent:#24364b;--accent-soft:#eef1f4;--border:#d8dde3;}
body{background:var(--bg);color:var(--ink);font-family:Inter,Segoe UI,Roboto,Arial,sans-serif;line-height:1.6;}
h1,h2,h3{font-weight:800;letter-spacing:-.01em;}
.card{background:#fff;border:1px solid var(--border);border-radius:14px;padding:18px 22px;margin:16px 0;box-shadow:0 7px 22px rgba(32,37,43,.04);}
.pill{display:inline-block;background:var(--accent-soft);color:var(--accent);border:1px solid var(--border);padding:5px 10px;border-radius:999px;font-weight:700;font-size:12px;text-transform:uppercase;letter-spacing:.04em;}
.note{background:#f4f6f7;border-left:4px solid var(--accent);padding:14px 16px;border-radius:8px;margin:14px 0;}
</style>

<div class="card">
<span class="pill">Applied seminar</span>
<h1>From MLE theory to market data</h1>
<p>This exercise applies the likelihood logic from the lecture to a simple financial-data example. We estimate the mean and volatility of daily S&amp;P 500 log returns under a Normal model.</p>
</div>

<div class="note"><strong>Modelling warning:</strong> the Normal model is a clean starting point, not a complete description of financial markets. The purpose is to learn how MLE is applied and then critique the assumptions.</div>

## Learning outcomes

By the end of this seminar, students should be able to:

1. load a financial time series in R;
2. convert prices into daily log returns;
3. write and evaluate a Normal log-likelihood;
4. estimate \(\mu\) and \(\sigma\) using closed-form MLE;
5. estimate the same parameters using `optim()`;
6. interpret and criticise the fitted model.

## 1. Setup

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

## 2. Load the S&P 500 data

The file `sp500_2017_fred.csv` should be in the same folder as this R Markdown file. If it is missing, the code tries to download the same data from FRED.

```{r load-data}
csv_file <- "sp500_2017_fred.csv"
fred_url <- "https://fred.stlouisfed.org/graph/fredgraph.csv?cosd=2017-01-01&coed=2017-12-31&id=SP500"

if (file.exists(csv_file)) {
  sp500_prices <- read.csv(csv_file)
} else {
  message("Local CSV not found. Attempting to download SP500 2017 data from FRED.")
  sp500_prices <- read.csv(fred_url)
}

if ("observation_date" %in% names(sp500_prices)) {
  names(sp500_prices)[names(sp500_prices) == "observation_date"] <- "Date"
}

head(sp500_prices)
```

## 3. Clean the data

We convert dates and prices into the correct formats, restrict attention to 2017, and remove missing price rows.

```{r clean-data}
sp500_prices$Date <- as.Date(sp500_prices$Date)
sp500_prices$SP500 <- as.numeric(sp500_prices$SP500)

sp500_prices <- sp500_prices[
  sp500_prices$Date >= as.Date("2017-01-01") &
    sp500_prices$Date <= as.Date("2017-12-31"),
]

sp500_clean <- sp500_prices[!is.na(sp500_prices$SP500), ]

nrow(sp500_clean)
head(sp500_clean)
```

## 4. Convert prices into log returns

We model returns rather than the raw index level. The daily log return is

\[
r_t=100	imes[\log(P_t)-\log(P_{t-1})].
\]

```{r returns}
returns <- 100 * diff(log(sp500_clean$SP500))
return_dates <- sp500_clean$Date[-1]

sp500_returns <- data.frame(
  Date = return_dates,
  Return = returns
)

length(returns)
head(sp500_returns)
```

## 5. Plot prices and returns

```{r plots, fig.width=9, fig.height=5}
plot(sp500_clean$Date, sp500_clean$SP500, type = "l",
     main = "S&P 500 daily close, 2017",
     xlab = "Date", ylab = "Index value")

plot(sp500_returns$Date, sp500_returns$Return, type = "l",
     main = "Daily percentage returns, 2017",
     xlab = "Date", ylab = "Daily percentage return (%)")
```

## 6. The Normal likelihood

Assume

\[
r_t\sim N(\mu,\sigma^2).
\]

For one return, the density is

\[
f(r_t\mid\mu,\sigma)=\frac{1}{\sigma\sqrt{2\pi}}\exp\left[-\frac{1}{2}\left(\frac{r_t-\mu}{\sigma}\right)^2\right].
\]

For all returns, the log-likelihood is

\[
\ell(\mu,\sigma)=\sum_{t=1}^T \log f(r_t\mid\mu,\sigma).
\]

## 7. Closed-form MLE

```{r closed-form-mle}
mu_hat <- mean(returns)
sigma_hat <- sqrt(mean((returns - mu_hat)^2))
annualised_volatility <- sigma_hat * sqrt(252)

mu_hat
sigma_hat
annualised_volatility
```

## 8. Numerical MLE using `optim()`

```{r numerical-mle}
negative_log_likelihood <- function(par, data) {
  mu <- par[1]
  sigma <- par[2]

  if (sigma <= 0) return(1e10)

  log_likelihood <- sum(dnorm(data, mean = mu, sd = sigma, log = TRUE))
  return(-log_likelihood)
}

mle_optim <- optim(
  par = c(mu = 0, sigma = 1),
  fn = negative_log_likelihood,
  data = returns
)

mle_optim$par
```

## 9. Compare formula and numerical estimates

```{r compare}
comparison <- data.frame(
  Method = c("Closed-form formula", "Numerical optimisation"),
  Mu = c(mu_hat, mle_optim$par["mu"]),
  Sigma = c(sigma_hat, mle_optim$par["sigma"])
)

comparison
```

## 10. Visual model check

```{r histogram-fit, fig.width=9, fig.height=5}
hist(returns, breaks = 25, probability = TRUE,
     main = "S&P 500 daily returns with fitted Normal density",
     xlab = "Daily percentage return (%)")

x_grid <- seq(min(returns), max(returns), length.out = 300)
fitted_density <- dnorm(x_grid, mean = mu_hat, sd = sigma_hat)
lines(x_grid, fitted_density, lwd = 2)
```

## 11. Discussion questions

1. Why do we model returns rather than prices?
2. What does \(\hat\mu\) mean in plain English?
3. What does \(\hat\sigma\) mean in plain English?
4. Why must \(\sigma\) be positive in the optimisation?
5. Where might the Normal model fail for financial returns?
